Fractional Types

### 2016-12-18 typespp.tex
move things around in sec 4
forall exists combined rule

@@

\begin{figure}[t]
{\footnotesize
\[\begin{array}{rrcll}
\unitetl/ :& [\order{\idiso} \circledast \mathbb{T},
                                                   \langle \idiso,v \rangle]
  & \leftrightblackspoon &
                   [\mathbb{T},v]
  &: \unititl/ \\
\unitetr/ :& [\mathbb{T} \circledast \order{\idiso},
                                                    \langle v,\idiso]
  & \leftrightblackspoon &
                   [\mathbb{T},v]
  &: \unititr/ \\
\swapt/ :& [\mathbb{T}_1 \circledast \mathbb{T}_2,
                                                   \langle v_1,v_2 \rangle]
 & \leftrightblackspoon &
                  [\mathbb{T}_2 \circledast \mathbb{T}_1,
                                                  \langle v_2,v_1 \rangle]
  &: \swapt/ \\
  \assoclt/ :& [\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
                                        \langle v_1,\langle v_2,v_3 \rangle]
  & \leftrightblackspoon &
                      [(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
                                       \langle \langle v_1,v_2 \rangle , v_3 \rangle]
  &: \assocrt/ \\
\\
\eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle] \\
\epsilon-: & \exists{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle]
  & \leftrightblackspoon & [\order{\idiso},\idiso] \\
\eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle] \\
\epsilon+ : & \exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle]
    & \leftrightblackspoon & [\order{\idiso},\idiso]
\end{array}\]
\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso/ :[\mathbb{T},v] \leftrightblackspoon
 [\mathbb{T},v]}}
{}
~~
\Rule{}
{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
{\jdg{}{}{(\order{\alpha})_k :
 [\order{c_1},c_1^k]  \leftrightblackspoon
 [\order{c_2}.c_2^k]}}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2]} \quad
  \rho_2 :
  [\mathbb{T}_2,v_2] \leftrightblackspoon [\mathbb{T}_3,v_3]}
{\jdg{}{}{\rho_1 \circledcirc \rho_2 :
 [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_3},v_3]}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
 [\mathbb{T}_1,v_1] \leftrightblackspoon
 [\mathbb{T}_2,v_2]}
 \qquad \rho_2 :
 [\mathbb{T}_3,v_3] \leftrightblackspoon
 [\mathbb{T}_4,v_4]}
{\jdg{}{}{\rho_1 \circledast \rho_2 :
  [\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
  \leftrightblackspoon
  [\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
{}
% \\
% \\
% \Rule{}
% {\jdg{}{}{}}
% {\jdg{}{}{\synch_i : \forall{\overline{k}k'}.[\mathbb{T},v]
%   \leftrightblackspoon
%    \forall{\overline{k}}.[\mathbb{T},v[k'\mapsto i]]}}
% {}
\end{array}\]
Each combinator $\rho_\bullet$ has an inverse.
}
\caption{$\Pi^/$ fractional combinators (dependent version)}
\label{pifracd:comb}
\end{figure}

@@

To understand the challenge in designing an operational semantics for $\eta$ and
$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
manipulation on types:
\[
x \mapsto (1 \times x) \mapsto ((x \times 1/x) \times x)
\mapsto (x \times (1/x \times x)) \mapsto (x \times 1) \mapsto x
\]
and by the coherence conditions of compact closed categories, we expect this
circuit, when run in either direction, to be equivalent to the
identity. Consider now the case of a combinator $c$ whose order is greater than
or equal to 2, i.e., the type $\order{c}$ has at least two distinct values
$\idiso$ and $c$ and the type $\iorder{c}$ has at least two distinct values
$1/\idiso$ and $1/c$. We require this circuit to produce $\idiso$ when given
$\idiso$ from either end, and similarly to produce $c$ when given $c$ from
either end. The problem becomes apparent when one observes that the uses of
$\eta+$ in the left-to-right execution and the use of $\epsilon-$ in the
right-to-left execution must behave differently depending on the incoming
value. But the ``local'' information available to either of these combinators
does not include the ``global'' information about the incoming value and there
is no way for them to make consistent guesses locally. It is possible to modify
the operational semantics to use a mechanism to express a computational effect
which enables spatially separated parts of the program to communicate in a way
that is reminiscent of entanglement in quantum mechanics. Possibilities to
realize this communication are global reference cells, backtracking, logical
variables with unification. We choose a more abstract approach that encodes the
required dependency in dataflow constraints using dependent types.

@@

The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
where each type $\mathbb{T}$ refers to a particular value $v$ in
$\mathbb{T}$. We then allow universal quantification over indices $k$
associated to values, over a collection of types.
In particular, the type
$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
imposes a constraint on the two values associated with the types $\order{c}$ and
$\iorder{c}$; they must be ``synchronized'' by sharing the same $k$. In
general, a pointed type will be of the form
$\forall\overline{k}.[\mathbb{T},v]$ for some sequence of integers
$\overline{k}$ and some value $v$ in $\mathbb{T}$, which might of course refer
to the integers of $\overline{k}$ to enforce some synchronization among its
sub-values.  The resulting combinators are shown in Fig.~\ref{pifrac:comb}.
We omit the quantifier from the notation when it is over the empty sequence
integers.  We recall that, as all fractional types are inhabited (see remark in
Sec.~\ref{sec:symmetryg}), and thus there is no loss in generality in going to
pointed types.

>>>>

To understand the challenge in designing an operational semantics for $\eta$ and
$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
manipulation on types:
\[
x \mapsto (1 \times x) \mapsto ((x \times 1/x) \times x)
\mapsto (x \times (1/x \times x)) \mapsto (x \times 1) \mapsto x
\]
and by the coherence conditions of compact closed categories, we expect this
circuit, when run in either direction, to be equivalent to the
identity. Consider now the case of a combinator $c$ whose order is greater than
or equal to 2, i.e., the type $\order{c}$ has at least two distinct values
$\idiso$ and $c$ and the type $\iorder{c}$ has at least two distinct values
$1/\idiso$ and $1/c$. We require this circuit to produce $\idiso$ when given
$\idiso$ from either end, and similarly to produce $c$ when given $c$ from
either end. The problem becomes apparent when one observes that the uses of
$\eta+$ in the left-to-right execution and the use of $\epsilon-$ in the
right-to-left execution must behave differently depending on the incoming
value. But the ``local'' information available to either of these combinators
does not include the ``global'' information about the incoming value and there
is no way for them to make consistent guesses locally. It is possible to modify
the operational semantics to use a mechanism to express a computational effect
which enables spatially separated parts of the program to communicate in a way
that is reminiscent of entanglement in quantum mechanics. Possibilities to
realize this communication are global reference cells, backtracking, logical
variables with unification. We present in the next section a more abstract
approach that encodes the required dependency in dataflow constraints using
dependent types.

@@

\begin{figure}[t]
{\footnotesize
\[\begin{array}{rrcll}
\unitetl/ :& [\order{\idiso} \circledast \mathbb{T},
                                                   \langle \idiso,v \rangle]
  & \leftrightblackspoon &
                   [\mathbb{T},v]
  &: \unititl/ \\
\unitetr/ :& [\mathbb{T} \circledast \order{\idiso},
                                                    \langle v,\idiso]
  & \leftrightblackspoon &
                   [\mathbb{T},v]
  &: \unititr/ \\
\swapt/ :& [\mathbb{T}_1 \circledast \mathbb{T}_2,
                                                   \langle v_1,v_2 \rangle]
 & \leftrightblackspoon &
                  [\mathbb{T}_2 \circledast \mathbb{T}_1,
                                                  \langle v_2,v_1 \rangle]
  &: \swapt/ \\
  \assoclt/ :& [\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
                                        \langle v_1,\langle v_2,v_3 \rangle]
  & \leftrightblackspoon &
                      [(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
                                       \langle \langle v_1,v_2 \rangle , v_3 \rangle]
  &: \assocrt/ \\
\\
\eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall\exists{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k  \rangle]
  & : \epsilon- \\
\eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall\exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle]
  & : \epsilon+
\end{array}\]
\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso/ :[\mathbb{T},v] \leftrightblackspoon
 [\mathbb{T},v]}}
{}
~~
\Rule{}
{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
{\jdg{}{}{(\order{\alpha})_k :
 [\order{c_1},c_1^k]  \leftrightblackspoon
 [\order{c_2}.c_2^k]}}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2]} \quad
  \rho_2 :
  [\mathbb{T}_2,v_2] \leftrightblackspoon [\mathbb{T}_3,v_3]}
{\jdg{}{}{\rho_1 \circledcirc \rho_2 :
 [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_3},v_3]}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
 [\mathbb{T}_1,v_1] \leftrightblackspoon
 [\mathbb{T}_2,v_2]}
 \qquad \rho_2 :
 [\mathbb{T}_3,v_3] \leftrightblackspoon
 [\mathbb{T}_4,v_4]}
{\jdg{}{}{\rho_1 \circledast \rho_2 :
  [\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
  \leftrightblackspoon
  [\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
{}
% \\
% \\
% \Rule{}
% {\jdg{}{}{}}
% {\jdg{}{}{\synch_i : \forall{\overline{k}k'}.[\mathbb{T},v]
%   \leftrightblackspoon
%    \forall{\overline{k}}.[\mathbb{T},v[k'\mapsto i]]}}
% {}
\end{array}\]
\\
Each combinator $\rho_\bullet$ has an inverse. A rule $\mathbb{T_1}_\bullet \leftrightblackspoon
\forall\exists{k}.\mathbb{T_2}_\bullet$ uses the $\forall$ in the left-to-right
direction and uses the $\exists$ in the right-to-left direction.
}
\caption{$\Pi^/$ fractional combinators (dependent version)}
\label{pifracd:comb}
\end{figure}

@@

The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
where each type $\mathbb{T}$ refers to a particular value $v$ in
$\mathbb{T}$. We recall that, as all fractional types are inhabited (see remark
in Sec.~\ref{sec:symmetryg}), and thus there is no loss in generality in going
to pointed types. We then allow universal quantification over indices $k$
associated to values, over a collection of types.  In particular, the type
$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
imposes a constraint on the two values associated with the types $\order{c}$ and
$\iorder{c}$; they must be ``synchronized'' by sharing the same $k$ but are
otherwise arbitrary. Once introduce a universal quantifier in one direction of
execution, we are led to also include an existential quantifier for the reverse
execution. Intuitively, a forward evaluation step that is at liberty to choose a
$k$ will, when reversed encounter a particular, i.e., existentially quantified,
$k$. The syntax is therefore as follows.

@@
